Query Optimization for Dynamic Imputation

### 2017-07-08 paper/algorithm.tex
paper: fix overfull hbox and length problems
paper: tighten up inline equations
paper fix: added $ to avoid latex compilation error in scalebox
paper: a pass over

\begin{align*}
  \textsc{Dirty}(q) = \begin{cases}
    \textsc{Dirty}(q') \setminus C & q = \delta_C(q')\ \text{or}\ q = \mu_C(q') \\
    \textsc{Dirty}(q') \cap C & q = \pi_C(q') \\
    \textsc{Dirty}(q_l) \cup \textsc{Dirty}(q_r) & q = q_l \bowtie_\psi q_r \\
    \textsc{Dirty}(q') & q = \sigma_\phi(q') \\
    \textsc{Dirty}(t) & q = \text{some table}\ t \\
  \end{cases}
\end{align*}

@@

\cameraready{
\begin{definition}
Let $Q$ be a set of plans. For $q \in Q$, we say $q$ is $\alpha$-bound if \[\textsc{Penalty}(q) - \min_{q' \in Q} \textsc{Penalty}(q') \leq \alpha.\] We denote this as $q^\alpha$.
Let the set of $\alpha$-bound plans in $Q$ be $Q^\alpha$.
We say a plan $q^\alpha$ is $\alpha$-bound optimal if \[\textsc{Time}(q^\alpha) = \min_{q' \in Q} \textsc{Time}(q').\]
\end{definition}
}

@@

Cardinality estimation for standard relational operators (\Crefrange{lst:std-start}{lst:std-end}) is the same as in a normal database.
\begin{algorithm}
\input{algorithms/histogram-transform.tex}
\end{algorithm}

\subsubsection{Imputation quality} 
\label{sec:quality}
Each imputation method has an associated cost function $P$, which is a measure of the quality of its output.
\ProjectName{} estimates the aggregate quality of the imputations in a plan $q$ using the heuristic measure $\textsc{Penalty}_{P}(q)$. \cameraready{Underlying
this is a helper function $\textsc{L}_{P(q)}$, which
is a recursive accumulator of penalties along imputation nodes instantiated
with an impute-specific penalty operator. This accumulator is computed
as follows.}

\begin{align*}
    &\textsc{L}_{P}(q) = \begin{cases}
     1 + \textsc{L}(q') & q = \delta_C(q') \\
     P(q') + \textsc{L}(q') & q = \mu_C(q') \\
     \textsc{L}(q_l') + \textsc{L}(q_r') & q = q_l' \bowtie_\psi q_r' \\
     \textsc{L}(q') & q = \sigma_\phi(q'),\ q = \pi_C(q') \\
     0 & q = \text{some table } t\\
  \end{cases}\\
    &\textsc{Penalty}_{P}(q) = \frac{\textsc{L}_{P}(q)}{\textsc{L}_{P_\delta}(q)}
\end{align*}

@@

\textbf{Constructing $P$}:
The form of $P$ is chosen to reflect the properties of the imputation method being used.
A well-chosen $P$ should provide a reasonable estimate of the quality of imputation from the set of plan histograms.
From these, $P$ can obtain the number of clean and dirty attributes in its input, their cardinalities, and approximations of distribution statistics such as mean and variance.
Intuitively, imputation quality \textit{increases} when more complete attributes and complete tuples are available; correspondingly, $P(q')$ \textit{decreases} as the number of values increases.
The penalty should also reflect the intuition that the number of training observations has diminishing returns by decreasing more slowly as the number of values increases.
Finally, the penalty may increase as amount of noise or spread in the data increases, as this makes it more difficult to obtain precise imputations.

@@

The form of $T$ depends on the specific imputation method, but there are several considerations in choosing $T$.
In particular, $T$ should \emph{increase} as the number of tuples and complete and dirty attributes passed to the imputation function increases and it should take into account the complexity of the imputation algorithm.
Analysis of the time complexity of different imputation algorithms, especially with in-database execution, is beyond the scope of this paper.
As with $P$, it is the ordering between queries generated by $T$ that is more important than the precision of the estimates.

@@

The chained-equation algorithm proceeds by
iteratively fitting decision trees for the target attributes. 
In each iteration, the missing values of a single attribute are replaced with newly imputed values.  Values in a single column can be updated over multiple
iterations of the algorithm. With more iterations, the quality of the
imputation improves as values progressively reflect more accurate relationships
between attributes. The algorithm terminates after a fixed number of
iterations, or earlier if convergence is achieved, indicated by unchanged values in a 
column in subsequent iterations. For our experiments, we use Weka's \texttt{REPTree}
decision tree implementation~\cite{witten2016data}, a variant of the canonical \texttt{C4.5}
algorithm~\cite{quinlan1993}. Roughly, the time complexity of building one
decision tree is $\mathcal{O}(nm \text{ } log(n))$, where $n$ is the number of tuples
and $m$ is the number of dependent attributes~\cite{witten2016data}. Since we build
$k * |\textsc{Dirty}(q)|$ trees, we have

\begin{equation*}
    \begin{aligned}
    T(q) = k * &|\textsc{Dirty}(q)| * \textsc{Card}(q) * \\
               &(|\textsc{Clean}(q)| + |\textsc{Dirty}(q)| - 1) * log(\textsc{Card}(q))
    \end{aligned}
\end{equation*}

Separately, we specify a heuristic penalty that takes into account the
improvement in imputation quality from more tuples and attributes, as well as
the diminishing returns from more data.

\[
    P(q) = \frac{|\textsc{Dirty}(q)|}{\sqrt{ \textsc{Attr}(q) * \textsc{Card}(q)}}
\]

@@

We can take advantage of the histograms maintained by \ProjectName{} to estimate the mean and construct a non-blocking impute operator.
Applying the heuristics from~\Cref{sec:quality} and~\Cref{sec:query-runtime}, we
have

\begin{equation*}
\begin{aligned}
    P(q) &= \sum_{c \in \textsc{Dirty}(q)} \frac{1+\widehat{Var}(c)}{\textsc{Card}(q)} \\
    T(q) &\approx 0
\end{aligned}
\end{equation*}

@@

Random hot deck imputation is a nonparametric method in which missing values in a column are
replaced with randomly selected complete values from the same column, preserving the
distribution of the complete values. We use the same $P$ as for mean value imputation, but
we modify $T$ to account for the fact that hot-deck is blocking in our implementation.

\begin{equation*}
\begin{aligned}
    P(q) &= \sum_{c \in \textsc{Dirty}(q)} \frac{1+\widehat{Var}(c)}{\textsc{Card}(q)} \\
    T(q) &= \textsc{Card}(q)
\end{aligned}
\end{equation*}
}

@@

\begin{figure}
  \cameraready{
  \begin{align*}
    \llbracket Q[T] \rrbracket &= \begin{cases}
      P & (T,\ P) \in Q \\
      \emptyset & \text{otherwise} \\
    \end{cases} \\
    \llbracket Q[T] \lhd p \rrbracket &= \begin{aligned} Q \cup \{(T,\ P) \} \end{aligned} \text{ where }\\
    P &= \left\{q \in S ~|~ \begin{aligned}
              \textsc{Dirty}(q) \neq \textsc{Dirty}(p) \lor
              \not \exists q' \in S:\ \doms{q'}{q}
              \end{aligned}
            \right\} \\
    S &= \llbracket Q[T] \rrbracket \cup \{p\}
  \end{align*}
  \caption{\cameraready{Semantics of a plan cache $Q$. $T$ is a set of tables and $p$ is a query plan. The cache guaranteees that plans for the same set of tables either have distinct dirty sets or are Pareto optimal.}}
  \label{fig:semantics-plan-cache}
  }
\end{figure}

@@

Finally, we note that for the queries we have examined, this exponential blowup does not affect the practical performance of our optimizer.
Recall that the planner maintains different plans for different dirty sets, keeping only those plans that are not dominated by others.
So in many cases we can drop some of the intermediate plans generated at each operator.
The worst case complexity only occurs if the dirty sets tracked are distinct through the entire planning phase.

>>>>

\begin{align*}
\scalebox{0.95}[1]{
  $\textsc{Dirty}(q) = \begin{cases}
    \textsc{Dirty}(q') \setminus C & q = \delta_C(q')\ \text{or}\ q = \mu_C(q') \\
    \textsc{Dirty}(q') \cap C & q = \pi_C(q') \\
    \textsc{Dirty}(q_l) \cup \textsc{Dirty}(q_r) & q = q_l \bowtie_\psi q_r \\
    \textsc{Dirty}(q') & q = \sigma_\phi(q') \\
    \textsc{Dirty}(t) & q = \text{some table}\ t \\
  \end{cases}$
}
\end{align*}

@@

\cameraready{
\begin{definition}
Let $Q$ be a set of plans. For $q \in Q$, we say $q$ is $\alpha$-bound if \[\textsc{Penalty}(q) - \min_{q' \in Q} \textsc{Penalty}(q') \leq \alpha.\] We denote this as $q^\alpha$.
Let the set of $\alpha$-bound plans in $Q$ be $Q^\alpha$.
We say a plan $q^\alpha$ is $\alpha$-bound optimal if \[\textsc{Time}(q^\alpha) = \min_{q' \in Q^\alpha} \textsc{Time}(q').\]
\end{definition}
}

@@

Cardinality estimation for standard relational operators (\Crefrange{lst:std-start}{lst:std-end}) is the same as in a typical database.
\begin{algorithm}
\input{algorithms/histogram-transform.tex}
\end{algorithm}

\subsubsection{Imputation quality} 
\label{sec:quality}
Each imputation method has an associated cost function $P$, which is a measure of the quality of its output.
\ProjectName{} estimates the aggregate quality of the imputations in a plan $q$ using the heuristic measure $\textsc{Penalty}_{P}(q)$. \cameraready{Underlying
this is a helper function, $\textsc{L}_{P(q)}$, which
is a recursive accumulator of penalties along imputation nodes instantiated
with an impute-specific penalty operator. This accumulator is computed
as follows.}
\begin{align*}
    &\textsc{L}_{P}(q) = \begin{cases}
     1 + \textsc{L}(q') & q = \delta_C(q') \\
     P(q') + \textsc{L}(q') & q = \mu_C(q') \\
     \textsc{L}(q_l') + \textsc{L}(q_r') & q = q_l' \bowtie_\psi q_r' \\
     \textsc{L}(q') & q = \sigma_\phi(q'),\ q = \pi_C(q') \\
     0 & q = \text{some table } t\\
  \end{cases}\\
    &\textsc{Penalty}_{P}(q) = \frac{\textsc{L}_{P}(q)}{\textsc{L}_{P_\delta}(q)}
\end{align*}

@@

\textbf{Constructing $P$}:
The form of $P$ is chosen to reflect the properties of the imputation method being used.
A well-chosen $P$ should provide a reasonable estimate of the quality of imputation from the set of plan histograms.
From these, $P$ can obtain the number of clean and dirty attributes in its input, their cardinalities, and approximations of distribution statistics such as mean and variance.
Intuitively, imputation quality \textit{increases} when more complete attributes and complete tuples are available; correspondingly, $P(q')$ \textit{decreases} as the number of values increases.
The penalty should also reflect the intuition that the number of training observations has diminishing returns by decreasing more slowly as the number of values increases.
Finally, the penalty may increase as the amount of noise or spread in the data increases, as this makes it more difficult to obtain precise imputations.

@@

The form of $T$ depends on the specific imputation method, but there are several considerations in choosing $T$.
In particular, $T$ should \emph{increase} as the number of tuples and complete and dirty attributes passed to the imputation function increases, and it should take into account the complexity of the imputation algorithm.
Analysis of the time complexity of different imputation algorithms, especially with in-database execution, is beyond the scope of this paper.
As with $P$, it is the ordering between queries imposed by $T$ that is more important than the precision of the estimates.

@@

The chained-equation algorithm proceeds by
iteratively fitting decision trees for the target attributes. 
In each iteration, the missing values of a single attribute are replaced with newly imputed values.  Values in a single column can be updated over multiple
iterations of the algorithm. With more iterations, the quality of the
imputation improves as values progressively reflect more accurate relationships
between attributes. The algorithm terminates after a fixed number of
iterations, or earlier if convergence is achieved, indicated by unchanged values in a 
column in subsequent iterations. For our experiments, we use Weka's \texttt{REPTree}
decision tree implementation~\cite{witten2016data}, a variant of the canonical \texttt{C4.5}
algorithm~\cite{quinlan1993}. Roughly, the time complexity of building one
decision tree is $\mathcal{O}(nm \text{ } log(n))$, where $n$ is the number of tuples
and $m$ is the number of dependent attributes~\cite{witten2016data}. Since we build
$k * |\textsc{Dirty}(q)|$ trees, we have
\begin{equation*}
    \begin{aligned}
    T(q) = k * &|\textsc{Dirty}(q)| * \textsc{Card}(q) * \\
               &(|\textsc{Clean}(q)| + |\textsc{Dirty}(q)| - 1) * log(\textsc{Card}(q)).
    \end{aligned}
\end{equation*}

Separately, we specify a heuristic penalty that takes into account the
improvement in imputation quality from more tuples and attributes, as well as
the diminishing returns from more data.
\[
    P(q) = \frac{|\textsc{Dirty}(q)|}{\sqrt{ \textsc{Attr}(q) * \textsc{Card}(q)}}
\]

@@

We can take advantage of the histograms maintained by \ProjectName{} to estimate the mean and construct a non-blocking impute operator.
Applying the heuristics from~\Cref{sec:quality} and~\Cref{sec:query-runtime}, we
have
\begin{equation*}
\begin{aligned}
    P(q) &= \sum_{c \in \textsc{Dirty}(q)} \frac{1+\widehat{Var}(c)}{\textsc{Card}(q)} \\
    T(q) &\approx 0.
\end{aligned}
\end{equation*}

@@

Random hot deck imputation is a non-parametric method in which missing values in a column are
replaced with randomly selected complete values from the same column, preserving the
distribution of the complete values. We use the same $P$ as for mean value imputation, but
we modify $T$ to account for the fact that hot-deck is blocking in our implementation.  
\begin{equation*}
\begin{aligned}
    P(q) &= \sum_{c \in \textsc{Dirty}(q)} \frac{1+\widehat{Var}(c)}{\textsc{Card}(q)} \\
    T(q) &= \textsc{Card}(q)
\end{aligned}
\end{equation*}
}

@@

\begin{figure}
  \cameraready{
  \begin{align*}
    \llbracket Q[T] \rrbracket &= \begin{cases}
      P & (T,\ P) \in Q \\
      \emptyset & \text{otherwise} \\
    \end{cases} \\
    \llbracket Q[T] \lhd p \rrbracket &= \begin{aligned} Q \cup \{(T,\ P) \} \end{aligned} \text{ where }\\
    P &= \left\{q \in S ~|~ \begin{aligned}
              \textsc{Dirty}(q) \neq \textsc{Dirty}(p) \lor
              \not \exists q' \in S:\ \doms{q'}{q}
              \end{aligned}
            \right\} \\
    S &= \llbracket Q[T] \rrbracket \cup \{p\}
  \end{align*}
  \caption{\cameraready{Semantics of a plan cache $Q$. We use $\llbracket\ \rrbracket$ to mean evaluation. $T$ is a set of tables and $p$ is a query plan. The cache guaranteees that plans for the same set of tables either have distinct dirty sets or are Pareto optimal.}}
  \label{fig:semantics-plan-cache}
  }
\end{figure}

@@

Finally, we note that for the queries we have examined, the exponential blowup does not affect the practical performance of our optimizer.
Recall that the planner maintains different plans for different dirty sets, keeping only those plans that are not dominated by others.
So in many cases we can drop some of the intermediate plans generated at each operator.
The worst case complexity only occurs if the dirty sets tracked are distinct through the entire planning phase.

### 2017-07-08 paper/introduction.tex
paper: add github url
paper: a pass over

% Solution
One common approach is to manually replace missing values using a statistical or predictive model based on other values in the table and record.
This process is called \emph{imputation}.
In this paper, we introduce new methods and theory to selectively applying
imputation to a subset of records, dynamically, during query execution.
We instantiate these ideas in a new system, \ProjectName{}.  \emph{The key insight behind \ProjectName{} is that imputation
only needs to be performed on the data relevant to a particular query and
that this subset is generally much smaller than the entire database.}  
% Background
While traditional imputation methods work over the entire data set and replace all missing
values, running a sophisticated imputation algorithm over a large data set can be very
expensive: our experiments show that even a relatively simple decision tree algorithm takes \acsbaseresulthours{} to train and run
on a 600K row database.

>>>>

% Solution
One common approach is to manually replace missing values using a statistical or predictive model based on other values in the table and record.
This process is called \emph{imputation}.
In this paper, we introduce new methods and theory to selectively apply
imputation to a subset of records, dynamically, during query execution.
We instantiate these ideas in a new system, \ProjectName{}\footnote{\ProjectUrl{}}.  \emph{The key insight behind \ProjectName{} is that imputation
only needs to be performed on the data relevant to a particular query and
that this subset is generally much smaller than the entire database.}  
% Background
While traditional imputation methods work over the entire data set and replace all missing
values, running a sophisticated imputation algorithm over a large data set can be very
expensive: our experiments show that even a relatively simple decision tree algorithm takes \acsbaseresulthours{} to train and run
on a 600K row database.
